\chapter{Software}
\label{chap:software}
\begin{overview}
  Even though solving a set of model equations by hand can also be
  seen as simulation~\citehere, modern simulation will almost certainly
  involve using computers.  The act of simulation (ie the actual
  solution of the modelling equations) is also not the only process
  which benefits by using computers.  Simulation generates data, which
  need to be stored and analysed in some way.  The model descriptions
  themselves need to be encoded and stored.  This chapter reviews
  simulation-related software in this larger context, discussing
  current technology for developing models, solving the model
  equations and storing the resulting data.  The choices that have
  been made for the modelling framework are interspersed in this
  discussion.
\end{overview}

\section{Model representations}

\subsection{Language based}

ProMOT \url{http://www.mpi-magdeburg.mpg.de/projects/promot}

Modelica

\subsection{XML based}

SBML \url{http://sbml.org/}


\subsection{Binary}


\section{Workflow}
NSF 

\subsection{What should workflow do?}

\subsection{Workflow tools}
% See:
% ~/Downloads/Documents/workflow
% ~/Downloads/Documents/workflow/scientific

\url{http://wiki.cogkit.org/index.php/Scientific_Workflow_Survey}

wftk \url{http://www.vivtek.com/wftk/perl_tutorial/08-workflow.html}
Perl-based single developer toolkit.  Good documentation about what
should go into a workflow tool.

Look into:
\begin{itemize}
\item \url{http://pegasus.isi.edu/wms}
\item \url{http://cppwfms.sourceforge.net}
\item \url{http://messagelab.monash.edu.au/Nimrod}
\end{itemize}

\subsection{Workflow specification languages}

\subsubsection{YAML}
\url{http://www.yawlfoundation.org} Very generic - examples are all
business workflows, has support for organisation model.  

\subsubsection{BPML}
Also for business.

\section{Concurrency}
\subsection{Local parallalism}

\section{Property packages}


\section{Simulators}
\subsection{Proprietary chemical systems}

\subsection{Modelica}
The Modelica project was spearheaded by \xxx, who started research on
modelling languages during his PhD studies in \xxx.  ``Modelica is a modern, strongly typed, declarative, and object-oriented language for modeling and simulation of complex systems.'' (\url{http://citeseerx.ist.psu.edu/viewdoc/summary?doi=10.1.1.139.7209})

\subsection{Stochastic simulation}

\subsection{Proprietary simulators}
\subsubsection{gPROMS}
\subsubsection{SPEEDUP}
\subsubsection{HYSIS}
\subsubsection{CHEMCAD}

\subsection{Zero cost}
\subsubsection{Model.la}
% Model.la_Jerry_Bieszczad_phd.pdf
By a student of Stephanopolous. \url{http://web.mit.edu/modella/}

\subsubsection{COCO}
COCO simulator - Based on the Cape Open standard.  \url{http://cocosimulator.org/}

\subsection{Open source}

\subsubsection{ASCEND}
ASCEND from Carnegie Mellon university.  From the ASCEND website: \url{http://ascend.cheme.cmu.edu/}
\begin{quote}
  ASCEND is a system for solving systems of equations, aimed at engineers and scientists. It allows you to build up complex models as as systems constructed from simpler sub-models. Using ASCEND it is simple to play around with your model, examine its behaviour, and work out how it can best be solved. You can easily change which variables are fixed and which are to be solved, and you can examine the way in which the model is being solved.
\end{quote}
\subsubsection{EMSO}
%TODO: Check out EML the EMSO model library
EMSO process simulator from ALSOC (\url{http://www.enq.ufrgs.br/trac/alsoc}).

From the EMSO website:
\begin{quote}
EMSO is the acronym for Environment for Modeling, Simulation, and Optimization. 
EMSO is a graphical environment where the user can model complex processes simply selecting and connecting the equipment models. 

The main features of EMSO follows: 

\begin{itemize}
\item Entirely written in C++
\item A fairly portable code, currently
  available for Windows and Linux but can be compiled for other
  platforms if desired
\item It is an Equation-Oriented simulator The unique
  Equation-Oriented simulator with units-of-measurement checking for
  the equations
\item A large set of built-in functions
\item Models are written
  in a modeling language, the user does not need to be a programmer

\item Models are converted to system of equations in memory, no
  compilation or linking is needed
\item An open library of models, called
  EML
\item Built-in code for symbolic differentiation which enables the
  system to solve high-index problems
\item Built-in code for automatic
  differentiation which makes the system very efficient
\item Can make use
  of machine optimize BLAS routines
\item Currently support: 
  \begin{itemize}
  \item static simulation
  \item dynamic simulation
  \item static optimization
  \item parameter estimation of static models
  \item parameter estimation of
    dynamic models
  \end{itemize}
\item A
  graphical user interface which can be used to model development,
  simulation execution, and results visualizing
\item A system of PlugIns where the user can embed code written in C,
  C++ or FORTRAN into the models A very modular system - all solvers
  are DLL's and the user can even write their own NewSolver Models
\end{itemize}
\end{quote}

\subsubsection{DWSIM}
\label{sec:dwsim}

From the DWSIM website: \url{http://dwsim.inforside.com.br/}
\begin{quote}
DWSIM is a Chemical Process Simulator for Windows. Built on the Microsoft .NET 2.0 Platform and featuring a rich Graphical User Interface (GUI), DWSIM allows chemical engineering students and chemical engineers to better understand the behavior of their chemical systems by using rigorous thermodynamic and unit operations' models with no cost at all. Even better, they can see how the calculations are actually being done - DWSIM is open source, that is, its code is available to anyone who wishes to discover the "magic" behind it or just do some code browsing. 
\end{quote}

\subsubsection{OPSIM}
\label{sec:opsim}
Windows-only.  \url{http://sourceforge.net/projects/opsim/}

\section{Programming languages}
\cite{chaves.nehrbass.ea2006octave}

\section{Data formats}
\subsection{Time-series data}
Input and output time-series

\subsection{Properties}
See CAPE-Open

\subsection{Object data}
Data requirements of each unit/storage

\section{Data technologies}
\subsection{Relational databases}
\subsection{HDF5}


% Local Variables:
% TeX-master: "thesis"
% End:

